Future cosmic microwave background delensing with galaxy surveys.

### 2017-10-29 Notes/galaxy_delensing.tex
closer to final

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, reducing the instrumental noise in future experiments will not be enough to detect this signal. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing part (``delensing'') from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in a multi-tracers approach to delensing that will improve the reconstruction of the lensing potential that lenses the CMB photons and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$). We find that, in general, galaxy surveys should be split into tomographic bins as this can improve the correlation with CMB lensing by 30\%. 
Among currently available surveys, a DES-like galaxy survey can remove about $14\%$ of the lensing signal. Ongoing or recently completed CMB experiments (CMB-S2) will particularly benefit from large scale structure tracers that, once properly combined, will have a better performance than a CMB internal reconstruction.
With the decrease of instrumental noise, the CMB internal reconstruction will increase its efficiency and the fraction of removed lensing B-modes with CMB alone will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) level.  
Nevertheless optical galaxy surveys will still play an important role even for CMB S4. In particular an LSST-like survey can a achieve a delensing performance comparable to a 3G CMB experiment but with completely different systematics. This will be important to prove the robustness against systematics of an eventual detection of primordial B-modes.
\end{abstract}

@@

The B-mode power spectrum measured on the sky is composed of a possible primordial component $C_{\ell}^{BB,r}$ together with the lensing $C_{\ell}^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_{\ell}^{BB}$ (defined in \refeq{noise}):
\be
C_{\ell}^{BB,\mathrm{measured}} = C_{\ell}^{BB,r} + C_{\ell}^{BB,\mathrm{lens}} + N_{\ell}^{BB}.
\ee
The lensing component is a significant source of B-modes that, at large scales, corresponds to a white noise source of roughly $5 \mu K$-arcmin independent of the angular scale. This means that it is not only larger than the allowed inflationary component at scales smaller than several degrees ($r_{0.05} < 0.12$ from \cite{bicep2/keck-collaboration:2015}), but it is also comparable to current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, while other approaches are possible (\cite{larsen:2016a,carron:2017,millea:2017}), here we assume a "template approach": we build a template \refeq{blens} of the lensing B-modes in the observed patch given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we can estimate $\phi$ using "tracers" of the matter distribution that sources the potential.

@@

%===========================
%===========================
\begin{table}
\caption{$\alpha(r)$: Improvements on $\sigma(r)$ due to delensing for completed and ongoing Stage-2 surveys and Planck lensing reconstruction.
LSS-S2 corresponds to the combination of all the available LSS tracers.
The values in parenthesis in the first column correspond to the fraction of lensing B-mode power removed using each LSS tracer. 
It is defined as $\left<C_{\ell}^{BB,\mathrm{res}}/C_{l}^{BB,\mathrm{lens}}\right>_{4<\ell<100}$.
%\newline
The values in the other columns correspond to ratio of the error before and after delensing for 3 cases: no instrumental noise in the B-mode measurement (but with instrumental noise in the E-mode used to delens) and with instrumental noise for two different values of $r$.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
WISE (8\%)& 1.08& 1.05 & 1.02 \\ \hline
DES (17\%) &1.20 & 1.13  & 1.05 \\ \hline
CIB (30\%) &1.44& 1.26  & 1.08\\ \hline
LSS-S2 (42\%) &1.75& 1.42  & 1.11\\ \hline
CMB Planck (7\%) &1.06 & 1.04  & 1.02 \\ \hline
LSS-S2 +CMB (45\%)& 1.81& 1.45& 1.13  \\ \hline
\end{tabular}
\label{tab:current}
\end{table}
%===========================
%===========================

We also test a scenario where primordial gravitational waves are present at their highest possible value of $r_{\rm{fid}}=0.12$ \cite{bicep2/keck-collaboration:2015}. As expected the importance of delensing itself is now reduced given that the lensing component constitutes a smaller portion of the total B-modes variance

 \begin{table}
\caption{$\alpha(r)$: Stage-2 improvements on $\sigma(r)$ due to delensing. We use the same LSS tracers as \reftab{current} but with the internal lensing reconstruction performed by SPTPol.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys &$\alpha_{r=0}, N^{B}_{\ell}=0$& $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
SPTPol (31\%) &1.43& 1.26  & 1.08 \\ \hline
LSS-S2 +SPTPol (55\%) & 2.2 & 1.6 & 1.14  \\ \hline

@@

We summarize the improvement in the constraint on $r$ in \reftab{S3}. 
Given the improvement in the noise of the high-resolution experiment, CMB alone will be able to improve constraint on $r$ by a factor of $\sim 2.4$ through internal delensing in the ideal case of no instrumental noise in the B-modes (but with instrumental noise in the E-mode used to delens).
Adding galaxy surveys will lead to a further improvement without additional effort.
Indeed galaxy surveys will still be able to remove an additional $11\%$ of power from $C_{\ell}^{BB,\mathrm{lens}}$ and to improve our constraint on the null hypothesis ($r=0$) by $34\%$. Note that apart from the addition of DESI, the tracers used here are already available today. The ongoing effort can then lead to a significant improvement even for the next generation of experiments. 
Finally, the combination of CMB and galaxy surveys will be able to improve the constrain on the null hypothesis of no primordial waves by a factor \textcolor{blue}{1.8} or to improve the constraint of a possible detection (r=0.07) by 34\%.
As expected the improvement on $n_{t}$ is similar to the one in $r$ since they are both proportional to the variance of the measured B-modes. 
We tested this in the scenario $r=0.12$ where we additionally impose the consistency relation $n_{t}^{\rm{fid}}=-r^{\rm{fid}}/8$. 
It is important to note that, independent of the level of delensing, for CMB 3G experiments the statistical error on $n_{t}$ will still be several times bigger than the fiducial value but it can still be tight enough to constrain more exotic inflationary models.

\begin{table}
\caption{$\alpha$: Improvements on $\sigma(r)$ due to delensing for S3 experiments. Here we define LSS-S3 as the combination of S2 LSS tracers (LSS-S2) and DESI. We also use a CMB S3 experiment for the lensing internal reconstruction. See \refsec{S3} for details.}
%The absolute error value of $\sigma(r)$ after delensing for $f_{sky} = $ is $\sigma(r) =$ 
\hspace{2.2cm}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
DESI (11\%) &1.13 & 1.08  & 1.03 \\ \hline
LSS-S3 (45\%) &1.82& 1.42  & 1.12\\ \hline
CMB 3G (56\%) &2.38& 1.7  & 1.16 \\ \hline
LSS-S3+CMB (68\%)& 3.2 & 2.3& 1.21  \\ \hline
\end{tabular}
\label{tab:S3}
\end{table}

@@

\begin{table}
\caption{$\alpha$: Improvements on $\sigma(r)$ due to delensing for S4 experiments. Here we define LSS-S4 as the combination of S3 LSS tracers (LSS-S3) and LSST. We also consider SKA-like radio-continuum surveys. We also use a CMB S4 experiment for the lensing internal reconstruction. See \refsec{S4} for details.}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
LSST (51\%) &2.0 & 1.9  & 1.7  \\ \hline
LSS-S4  (59\%) &2.5 & 2.3  & 2.3 \\ \hline
SKA (10$\mu$Jy) (53\%) &2.1& 1.9  & 1.16\\ \hline
LSS-S4 + SKA (10$\mu$Jy)  (62\%) &2.6 & 2.3  & 2.5  \\ \hline
CMB S4 (81\%) & 5.0 & 3.7  & 4.2 \\ \hline
LSS-S4+CMB (86\%)& 6.6 & 4.3  &  \\ \hline
\end{tabular}
\label{tab:S4}
\end{table}

@@

\section{Conclusions}
\label{sec:concl}
The ability to separate the lensing component of the CMB B-modes from a possible primordial inflationary signal (``delensing'') is necessary to test inflation using the next generation of CMB polarization experiments. 
To delens, we need to accurately reconstruct the large scale structures that lens the CMB in order to build a template of the expected lensing B-modes in the observed patch.
In this paper, we studied the potential impact of large-scale structure galaxy surveys in this important endeavor. 
We find that, to improve the delensing efficiency of galaxy surveys, tomographic delensing will be very important: using several tomographic galaxy bins can improve the correlation of galaxies with CMB lensing by 10\%-30\% on a wide range of angular scales.

>>>>

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, reducing the instrumental noise in future experiments will not be enough to detect this signal. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing part (``delensing'') from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in a multi-tracers approach to delensing that will improve the reconstruction of the lensing potential that lenses the CMB photons and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$). We find that, in general, galaxy surveys should be split into tomographic bins as this can improve the correlation with CMB lensing by 30\%. 
Among currently available surveys, a DES-like galaxy survey can remove about $17\%$ of the lensing signal. Ongoing or recently completed CMB experiments (CMB-S2) will particularly benefit from large scale structure tracers that, once properly combined, will have a better performance than a CMB internal reconstruction.
With the decrease of instrumental noise, the CMB internal reconstruction will increase its efficiency and the fraction of removed lensing B-modes with CMB alone will rapidly improve from the current level of Planck ($7\%$) and SPTPol ($31\%$) to 3G (56\%) and CMB S4 (81\%) level.  
Nevertheless optical galaxy surveys will still play an important role even for CMB S4. In particular an LSST-like survey can a achieve a delensing performance comparable to a 3G CMB experiment but with completely different systematics. This will be important to prove the robustness against systematics of an eventual detection of primordial B-modes.
\end{abstract}

@@

The B-mode power spectrum measured on the sky is composed of a possible primordial component $C_{\ell}^{BB,r}$ together with the lensing $C_{\ell}^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_{\ell}^{BB}$ (defined in \refeq{noise}):
\be
C_{\ell}^{BB,\mathrm{measured}} = C_{\ell}^{BB,r} + C_{\ell}^{BB,\mathrm{lens}} + N_{\ell}^{BB}.
\ee
The lensing component is a significant source of B-modes that, at large scales, corresponds to a white noise source of roughly $5 \mu K$-arcmin independent of the angular scale. This means that it is not only larger than the allowed inflationary component at scales smaller than several degrees ($r_{0.05} < 0.07$ from \cite{bicep2-collaboration:2016a}), but it is also comparable to current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, while other approaches are possible (\cite{larsen:2016a,carron:2017,millea:2017}), here we assume a "template approach": we build a template \refeq{blens} of the lensing B-modes in the observed patch given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we can estimate $\phi$ using "tracers" of the matter distribution that sources the potential.

@@

%===========================
%===========================
\begin{table}
\caption{$\alpha(r)$: Improvements on $\sigma(r)$ due to delensing for completed and ongoing Stage-2 surveys and Planck lensing reconstruction.
LSS-S2 corresponds to the combination of all the available LSS tracers.
The values in parenthesis in the first column correspond to the fraction of lensing B-mode power removed using each LSS tracer. 
It is defined as $\left<C_{\ell}^{BB,\mathrm{res}}/C_{l}^{BB,\mathrm{lens}}\right>_{4<\ell<100}$.
%\newline
The values in the other columns correspond to ratio of the error before and after delensing for 3 cases: no instrumental noise in the B-mode measurement (but with instrumental noise in the E-mode used to delens) and with instrumental noise for two different values of $r$.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.07}$ \\ \hline \hline
WISE (8\%)& 1.08& 1.05 & 1.02 \\ \hline
DES (17\%) &1.20 & 1.13  & 1.05 \\ \hline
CIB (30\%) &1.44& 1.26  & 1.10\\ \hline
LSS-S2 (42\%) &1.75& 1.42  & 1.15\\ \hline
CMB Planck (7\%) &1.06 & 1.04  & 1.02 \\ \hline
LSS-S2 +CMB (45\%)& 1.81& 1.45& 1.16  \\ \hline
\end{tabular}
\label{tab:current}
\end{table}
%===========================
%===========================

We also test a scenario where primordial gravitational waves are present at their highest possible value of $r_{\rm{fid}}=0.07$ \cite{bicep2-collaboration:2016a}. As expected the importance of delensing itself is now reduced given that the lensing component constitutes a smaller portion of the total B-modes variance

 \begin{table}
\caption{$\alpha(r)$: Stage-2 improvements on $\sigma(r)$ due to delensing. We use the same LSS tracers as \reftab{current} but with the internal lensing reconstruction performed by SPTPol.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys &$\alpha_{r=0}, N^{B}_{\ell}=0$& $\alpha_{r=0}$ & $\alpha_{r=0.07}$ \\ \hline \hline
SPTPol (31\%) &1.43& 1.26  & 1.10 \\ \hline
LSS-S2 +SPTPol (55\%) & 2.2 & 1.6 & 1.20  \\ \hline

@@

We summarize the improvement in the constraint on $r$ in \reftab{S3}. 
Given the improvement in the noise of the high-resolution experiment, CMB alone will be able to improve constraint on $r$ by a factor of $\sim 2.4$ through internal delensing in the ideal case of no instrumental noise in the B-modes (but with instrumental noise in the E-mode used to delens).
Adding galaxy surveys will lead to a further improvement without additional effort.
Indeed galaxy surveys will still be able to remove an additional $11\%$ of power from $C_{\ell}^{BB,\mathrm{lens}}$ and to improve our constraint on the null hypothesis ($r=0$) by $34\%$. Note that apart from the addition of DESI, the tracers used here are already available today. The ongoing effort can then lead to a significant improvement even for the next generation of experiments. 
Finally, the combination of CMB and galaxy surveys will be able to improve the constrain on the null hypothesis of no primordial waves by a factor \textcolor{blue}{1.8} or to improve the constraint of a possible detection (r=0.07) by 34\%.
As expected the improvement on $n_{t}$ is similar to the one in $r$ since they are both proportional to the variance of the measured B-modes. 
We tested this in the scenario $r=0.07$ where we additionally impose the consistency relation $n_{t}^{\rm{fid}}=-r^{\rm{fid}}/8$. 
It is important to note that, independent of the level of delensing, for CMB 3G experiments the statistical error on $n_{t}$ will still be several times bigger than the fiducial value but it can still be tight enough to constrain more exotic inflationary models.

\begin{table}
\caption{$\alpha$: Improvements on $\sigma(r)$ due to delensing for S3 experiments. Here we define LSS-S3 as the combination of S2 LSS tracers (LSS-S2) and DESI. We also use a CMB S3 experiment for the lensing internal reconstruction. See \refsec{S3} for details.}
%The absolute error value of $\sigma(r)$ after delensing for $f_{sky} = $ is $\sigma(r) =$ 
\hspace{2.2cm}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.07}$ \\ \hline \hline
DESI (11\%) &1.13 & 1.08  & 1.03 \\ \hline
LSS-S3 (45\%) &1.82& 1.59  & 1.18\\ \hline
CMB 3G (56\%) &2.31& 1.65  & 1.20 \\ \hline
LSS-S3+CMB (68\%)& 3.2 & 2.3& 1.21  \\ \hline
\end{tabular}
\label{tab:S3}
\end{table}

@@

\begin{table}
\caption{$\alpha$: Improvements on $\sigma(r)$ due to delensing for S4 experiments. Here we define LSS-S4 as the combination of S3 LSS tracers (LSS-S3) and LSST. We also consider SKA-like radio-continuum surveys. We also use a CMB S4 experiment for the lensing internal reconstruction. See \refsec{S4} for details.}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.07}$ \\ \hline \hline
LSST (51\%) &2.0 & 1.9  & 1.21  \\ \hline
LSS-S4  (59\%) &2.4 & 2.1  & 1.25 \\ \hline
SKA (10$\mu$Jy) (53\%) &2.1& 1.9  & 1.22\\ \hline
LSS-S4 + SKA (10$\mu$Jy)  (62\%) &2.6 & 2.3  & 2.5  \\ \hline
CMB S4 (81\%) & 5.0 & 3.7  & 1.43 \\ \hline
LSS-S4+CMB (86\%)& 6.6 & 4.3  &  \\ \hline
\end{tabular}
\label{tab:S4}
\end{table}

@@

\section{Conclusions}
\label{sec:concl}
The ability to separate the lensing component of the CMB B-modes from a possible primordial inflationary signal (``delensing'') is necessary to test inflation using the next generation of CMB polarization experiments. 
To delens, we need to accurately reconstruct the large scale structures that lens the CMB in order to build a template of the expected lensing B-modes in the observed patch.
In this paper, we studied the potential impact of large-scale structure galaxy surveys in this important endeavor. 
We focus on galaxy clustering here, leaving other probes like weak lensing, for future works. However,  weak lensing is usually lower redshift probes than clustering resulting in a worse delensing efficiency. 
We find that, to improve the delensing efficiency of galaxy surveys, tomographic delensing will be very important: using several tomographic galaxy bins can improve the correlation of galaxies with CMB lensing by 10\%-30\% on a wide range of angular scales.
